--R
title: "Machine Learning 1 - Regression"
output: html_notebook
---

## Linear Regression

Use the built-in dataset "cars" to find a regression relating speed and distance

```{r}
str(cars)
scatter.smooth(x=cars$speed,y=cars$dist, main="Dist ~ Speed")

Reg <- lm(dist ~ speed, data=cars)
summary(Reg)

```
 
Multiple Regression
using 


model <- lm(Dependent variable ~ First independent Variable + Second independent variable + ...)


```{r}
Year <- c(2017,2017,2017,2017,2017,2017,2017,2017,2017,2017,2017,2017,2016,2016,2016,2016,2016,2016,2016,2016,2016,2016,2016,2016)
Month <- c(12, 11,10,9,8,7,6,5,4,3,2,1,12,11,10,9,8,7,6,5,4,3,2,1)
Interest_Rate <- c(2.75,2.5,2.5,2.5,2.5,2.5,2.5,2.25,2.25,2.25,2,2,2,1.75,1.75,1.75,1.75,1.75,1.75,1.75,1.75,1.75,1.75,1.75)
Unemployment_Rate <- c(5.3,5.3,5.3,5.3,5.4,5.6,5.5,5.5,5.5,5.6,5.7,5.9,6,5.9,5.8,6.1,6.2,6.1,6.1,6.1,5.9,6.2,6.2,6.1)
Stock_Index_Price <- c(1464,1394,1357,1293,1256,1254,1234,1195,1159,1167,1130,1075,1047,965,943,958,971,949,884,866,876,822,704,719)        
                
model <- lm(Stock_Index_Price ~ Interest_Rate + Unemployment_Rate)
summary(model)
```

## Logistic Regression
example is at https://towardsdatascience.com/machine-learning-with-r-logistic-regression-152ec20351db

```{r}
library(dplyr)
library(stringr)
install.packages('caTools')
install.packages('caret')
library(caTools)
library(caret)

d1f <- read.csv('titanic.csv')
str(df)

```

There are a couple of essential things we have to do:

    Extract titles from the Name attribute
    Remap extracted titles as usual/unusual
    Convert Cabin attribute to binary — HasCabin
    Remove unnecessary attributes
    
```{r}
maleNobleTitles <- c('Capt', 'Col', 'Don', 'Dr', 'Jonkheer', 'Major', 'Rev', 'Sir')

femaleNobleTitles <- c('Lady', 'Mlle', 'Mme', 'Ms', 'the Countess') 
df$Title <- str_sub(df$Name, str_locate(df$Name, ',')[ , 1] + 2, str_locate(df$Name, '\\.')[ , 1] - 1) 

df$Title[df$Title %in% maleNobleTitles] <- 'MaleNoble' 
df$Title[df$Title %in% femaleNobleTitles] <- 'FemaleNoble' 
df$HasCabin <- ifelse(df$Cabin == '', 0, 1) 
df <- df %>% select(-PassengerId, -Name, -Ticket, -Cabin)
```
Transform data replace missing ages with average age

```{R}

lapply(df, function(x) { length(which(is.na(x))) })
df$Age <- ifelse(is.na(df$Age), mean(df$Age, na.rm=TRUE), df$Age)

```

Convert categorical data to factors

```{r}
df$Survived <- factor(df$Survived) 
df$Pclass <- factor(df$Pclass) 
df$Sex <- factor(df$Sex) 
df$SibSp <- factor(df$SibSp) 
df$Parch <- factor(df$Parch) 
df$Embarked <- factor(df$Embarked) 
df$Title <- factor(df$Title) 
df$HasCabin <- factor(df$HasCabin)

```






