---
title: "Airline Analysis"
output: html_notebook
---
## Project Description

Comparative analysis of airline delay data to determine:

1. What factor causes the most number of delays in terms of both flights and time
2. Amount each factor contributes to overall delays
3. Worst airports for delays overall and factor by each airport that causes the most delays
4. Causes and amount of delays by year and by month
5. Delays for each airport by year

### Data Sources

Data is sourced from:

https://corgis-edu.github.io/corgis/csv/airlines/

### Data Aquisition

Read in data from csv file
```{r}
df <- read.csv("airlines.csv", stringsAsFactors = FALSE)
```

### Perform some preliminary analysis

Run some basic analysis to see the shape of the data
```{r}
str(df)
dim(df)
ncol(df)
nrow(df)
summary(df)
colnames(df)
rownames(df)
length(unique(df[,1]))
```

### Data Transformation

Each of the questions above looks at a only a subset of the data. A first step is to figure out how to extract the data we are interested in so that we don't process a lot of data we don't need.

For this example, we will look at the overall delays by flights and by minutes. In this case, a number f columns are redundant so we can eliminate the Airport.Names, Time.Label, Time.Month.Name and the list of airlines and number of airlines.
```{r}
df1 <- subset(df,select=-c(2,3,5,12,13) )
colnames(df1)

```

The Column names are awkward, so rename columns with a new length(19) character vector. 
```{r}
 newnames <-  c("Code", "Month", "Year" , "Carriers", "LatePlane", "NVS", "Security", "Weather", "Cancelled", "Delayed", "Diverted", "OnTime", "Total", "MinCarrier", "MinLateAir", "MinNVS", "MinSec", "MinTotal", "MinWeather")
colnames(df1) <- newnames
```

We also want to reverse the order of the last two columns
```{r}
df2 <- cbind(df1[,1:17],df1[,19],df1[,18])
colnames(df2)
```

But the last two columns lose their names so we rename
```{r}
 newnames <-  c("Code", "Month", "Year" , "Carriers", "LatePlane", "NVS", "Security", "Weather", "Cancelled", "Delayed", "Diverted", "OnTime", "Total", "MinCarrier", "MinLateAir", "MinNVS", "MinSec", "MinWeather", "MinTotal")
colnames(df2) <- newnames
```

#### Data Quality

Not sure if the columns were reversed correctly so compare the columns from before and after to make sure they are correct
```{r}
 ctotal <- df1$MinTotal == df2$MinTotal
 length(ctotal[ctotal == FALSE])
 cweather <- df1$MinWeather == df2$MinWeather
 length(cweather[cweather == FALSE])
 
```

### Data Transformation.

For our first analysis, suppose we are only interested in totals over years in question. If we collapse the data by airport, how many rows would that produce?
```{r}
length(unique(df2$Code))
```

So we have 29 airports. So now we get rid of the columns that have to do with years and months.
```{r}
df3 <- subset(df2, select=-c(2,3))
```

But we need to aggregate the data by airport. Since we are collapsing over the years, we should have the same number of observations for each airport.
```{r}
aggregate.data.frame(df3[,1], list(df3$Code), length)
df4 <- aggregate.data.frame(df3[,2:17], list(df3$Code), sum)

```

If we want to work with just the number of flights:
```{r}
fl <- subset(df4, select=c(1:10))
```

Or if we want to work with minutes
```{r}
mts <- subset(df4, select=c(1,12:17))
```

At this point, we know the steps that we need to do to extract this data for analysis, so we can combine the steps into a function called "extractFlightDelaysByAirport()" and another function called "extractAirportDelaysByMinutes()"
```{r}
## To come
```

## Data Visualization

We can use some basic plotting functions to start to visualise the data
For example, Boxplot
```{r}
boxplot(mts$MinCarrier,mts$MinTotal)
```

But these numbers are hard to read so look at percentages instead
```{r}
boxplot((mts$MinCarrier/mts$MinTotal)*100,(mts$MinWeather/mts$MinTotal)*100)
```

This suggests we can transform the raw data into a more workable form of percentages for some analyses
```{r}
permts <- data.frame(mts$Group.1,100*(mts[,2:7]/mts$MinTotal))

```

However, the raw data may still be useful for other kinds of analyses like the ordering of the most number of minutes

```{r}
mts[order(-mts$MinTotal),]
```

we can also extract data by cities

```{r}
atl <- subset(df,Airport.Code == "ATL")

```

And we can save our datasets
```{r}
save(mts,file="mts.Rda")
write.csv(mts,"mts.csv")
```

