---
title: "Welcome to the tidyverse"
output: html_notebook
---

### Installing some packages.

The devtools package allows for compiling packages for efficiency

```{r}
install.packages("devtools")
install.packages("tidyverse")
install.packages("nycflights13")
devtools::install_github("rstudio/EDAWR")
library(EDAWR)
library(nycflights13)
library(tidyverse)

```

Examine some data
Note that they are all different in logical structure


```{r}
storms
?storms

cases
?cases

pollution
?pollution


```

The storms df is easier to use because we have variables as columns. Consider cases, we really have two variables

```{r}
cases$country
names(cases[-1])
unlist(cases[1:3, 2:4])

```

## Defining Tidy Data

Each Variable is in its own column
Each Observation is in its own row
Each type of data is in its own "table"

Two main functions: gather() and spread()
```{r}
gather(cases, "year", "n", 2:4)
```

Consider the pollution dataset

```{r}
spread(pollution, size,amount)
```

Consider storms, the date column is really three different variables
We can collapse and separate columns using string operations

```{r}
?storms
x <- EDAWR::storms
x1 <- separate(x,date,c("year","month","day"),sep="-")
unite(x1,"date",year,month,day,sep="/")
```
===========================

# Dplyr Working with data


```{r}
?select
?filter
?arrange
?mutate
?summarise
?group_by
```

Using flights data

```{r}
library(nycflights13)
?flights
?airports
?airlines
?planes
?weather
```


Operations

Extract columns or variables
Extract rows or observations
Derive new variables
Transform the values of observations

```{r}
x
select(x,storm,pressure)
select(x,-pressure)
```
Multiple ways to select

```{r}
select(x,starts_with("w"))
select(x, contains("re"))
select(x,everything())

```

Filter selects rows based on criteria

```{r}
filter(x, wind > 50)
filter(x, storm %in% c("Alberto","Alex","Allison"))
filter(x, storm %in% c("Alberto","Alex","Allison"),wind > 50)
filter(x, storm %in% c("Alberto","Alex","Allison") | wind < 50)
```

Mutate creates new variables. There are a number of helper functions we can use as well

```{r}
mutate(x, ratio = pressure / wind)
```

Summarise works like the summary function we saw before

```{r}
summarise(pollution, median=median(amount),mean=mean(amount),variance = var(amount))

```

We can also use a more functional programming syntax using a "pipe"
ctl-shift-M inserts the operator
```{r}
pollution %>% summarise(median=median(amount),mean=mean(amount),variance = var(amount))

```

Arrange sorts the data

```{r}
x %>% arrange(wind)
arrange(x,-wind)
```
We can sort by two variables using the second as a tie-breaker
```{r}
arrange(x, wind, date)
```
We can easily combine expressions using pipes
```{r}
x %>% filter(wind > 50) %>% select(storm,wind)
```

Group by.  Allows us to group observations by asdding metadata to the dataframe

```{r}
p<-pollution
group_by(p,size) 
p %>% group_by(city) %>% summarise(mean=mean(amount),sum=sum(amount),n=n())
p %>% group_by(size) %>% summarise(mean=mean(amount))
```


