---
title: "Object Oriented Programming"
output: html_notebook
---

# OOP 

OOP is designed to emulate objects in the real world and their interactions

An object has some conceptual reality and is of some type. For example, "Person" is a type.

Objects are instances of "types".  Rod is an object which is an instance of "Person"

Objects can be instances of several types: For example I am a person, a parent and a Canadian

All objects of a given type are assumed to have the same properties and attributes.

Types define interfaces (how we interact with the object) and how it interacts with objects of other types
  - For example, how an object of type "teacher" interacts with an object of type "student"
  
In OOP types define:
  1. Attribute:, the data associated with an object. For a person object this might be:
    - height, weight, date of birth, etc
  2. Behaviours - actions object perform when receiving a message.
    - Polymorphism says that what an object does when it receives a message depends on its type
    - for example "Fire!" will be handled differently by a firefighter object than a soldier object.
  3. Identity - some way of uniquely identifying an object from all other objects of that type
  
OOP is iconic

  1. We think in terms of types and object when dealing with the world around us
  2. OOP says programs should look like what they automate.
    Users of the program should see the same objects when they interact with an application that they understand from their ral world experience.
  
For data analysis, looking at the world in terms of objects can be most useful since we are capturing data about objects in the real world.
  1. For example a survey of voting preferences would naturally use "person" as an object
    -leads to natural analysis like "How likely is it that a person who is male and lives in the suburbs would vote for Doug Ford?"
    
## OOP in R

R was not intended as an OOP language but has incorporated OOP capabilities

- S3 and S4 are OOP capabilities that are not part of base R
- R6 is an OO package that 
- R had a base object system before it adopted OO so it can be a bit confusing.
- the main difference is that base R objects do not have a type associated with them
  - base objects return FALSE when queried if they are objects
- Every object has a base type even if it is an OO object
  - These are not "object oriented" types but work at the underlying C code leve
  - The terminology is confusing because of the historically different uses of the terms type and class
  
  S3, S4, R6
  

```{r}

ls <- list()
is.object(ls)
attr(ls,"class")
typeof(ls)


df <- data.frame()
is.object(df)
attr(df,"class")

install.packages('sloop')
sloop::otype(ls)
sloop::otype(df)
typeof(df)

```

## S3

An S3 object is an instance of some class and has optional attributes

For example, a student type is defined to have attributes of name, age and GPA  

```{r}

s <- list(Name = "John", Age = 21, GPA = 3.5, StudentID=99879)
class(s) <- "Student"

is.object(s)
sloop::otype(s)
attr(s,"class")

```

Best practices is to provide three functions:
  1. A constructor to ensure that an object is structurally correct (Ensure consistent naming of fields for example)
  2. A Validator to ensure that the object is a correct state (the age is non-zero for example)
  3. A factory function that can be called to create new objects (often will call the constructor and validator)
    - Also may ensure that only one object - for example, if a new student is requested to be created with a given student number, but the number is un use, the factory might return a reference to the existing student object (remember identity)
    
```{r}

newStudent <- function(num,nam,age,gpa) {
  
  structure(list(Name = nam, Age = age, GPA = gpa, StudentID=num), class="Student")
}
  b <- newStudent(34478,"Bill",21, 3.5)
  attr(b,"class")

studentValidator <- function(age,gpa) {
  if (!is.numeric(gpa) || gpa <=0 || gpa > 4) return(FALSE)
  if (!is.numeric(age) || age <=0) return(FALSE)
  return(TRUE)
}


newStudent2 <- function(num,nam,age,gpa) {
  
  if(!studentValidator(age,gpa)) {
    stop("Invalid data supplied for student object", call. = TRUE)
  }
  
  structure(list(Name = nam, Age = age, GPA = gpa, StudentID=num), class="student")
}
```

Factory methods are used to call the constructor and validator and do other work.
For example, enforce the singleton pattern

let us suppose that there can be only one student object (for whatever reason)

```{r}

Singleton <- NULL

Student <-  function(num,nam,age,gpa) {
  if (is.null(Singleton)) Singleton <<- newStudent2(num,nam,age,gpa)
  return (Singleton)
}

Student(34478,"Bill",21, 3.5)
```

## Writing Methods

A method like print is implemented differently for different kinds of objects. 
When using print with an object the correct version is "dispatched" or executed depending on the class

For example, to write a print method for a student. The print() function is polymorphic in that it has different bodies which are executed depending on the type of thing it is printing. We can add to the list of specializations like adding a print method for a Student object

```{r}
print.Student <- function(obj) {
cat(obj$Name, " ID: ", obj$StudentID,"\n")
cat(obj$Age, "years old\n")
cat("GPA:", obj$GPA, "\n")
}
```

One of the advantages to OOP is that we can let the object containing the data decide how to process a function.
This means we can have a collection of objects, each of which has a different implementation of a method

Then if we map the generic function, 

```{r}
d <- list(Name = "Dog")
class(d) <- "Dog"
b <- list(Name = "Bird")
class(b) <- "Bird"

print.Dog <- function(obj) cat("Arf \n")
print.Bird <- function(obj) cat("Tweet \n")

zoo <- list(d,b,bill)

sooout <-map(zoo,print)
```

Or we can create our own generic

```{r}
speak <- function(obj) {
  UseMethod("speak")
}

speak.default <- function(obj) ".... crickets ...."
speak.Bird <- function(obj) "Melodic Song"
speak.Dog <- function(obj) "Arf arf"


s <-  map(zoo,speak)
```
-------------

# s4

S4 is a more formal OO programming style

You define an S4 class by calling setClass() with the class name and a definition of its slots, and the names and classes of the class data:

```{r}

library(methods)

setClass("Person", 
  slots = c(
    name = "character", 
    age = "numeric"
  )
)

john <- new("Person", name = "John Smith", age = NA_real_)
is(john)
john@name
slot(john, "name")

sloop::otype(john)

```

Getter and Setter functions. We need a base template to work from


```{r}
setGeneric("age", function(x) standardGeneric("age"))
setGeneric("age<-", function(x, value) standardGeneric("age<-"))


```

The getters and setters ensure that the data is accessed only through the interface of the object

```{r}
setMethod("age", "Person", function(x) x@age)
setMethod("age<-", "Person", function(x, value) {
  x@age <- value
  x
})

age(john) <- 50
age(john)

sloop::otype(john)
sloop::ftype(age)
```


Using Prototypes

Prototypes can be used in define default structure

```{r}

setClass("Car",
    slots = c(
        cylinders = "numeric",
        maker = "character"
     ),
     prototype = list(
         cylinders = NA_integer_,
           maker = NA_character_
      )
)

Fiesta <- new("Car", cylinders = 4, maker = "Ford")

str(Fiesta)
```

Inheritance allows us to build specializations

```{r}
setClass("Sedan",
  contains = "Car",
  slots = c( doors = "numeric",
             owner = "Person"),
  prototype = list (
    doors = 4,
    owner = new("Person")
  ))

Acura <- new("Sedan")
str(Acura)
is(Acura, "Car")
```

Constructors and Validators

new is used as a constructor, so we want to wrap it in a factory method which, by convention is the same as the class name

```{r}
Person <- function(name, age = NA) {
  age <- as.double(age)
  
  new("Person", name = name, age = age)
}

Person("Plato")
```

Validators do the same as in S3

```{r}
Person(45)  # This is an error

setValidity("Person", function(object) {
  if (!is.na(object@age) && object@age <= 0) {
    "@age must be greater than zero"
  } else {
    TRUE
  }
})

Person("Zemo", -1)
```

1. Override - replace base class method with derived class method
2. Extension - execute base class method but also do addition subclass stuff

Things can get really complicated really fast.

1. Design - think through the design
2. Don't write classes "just because" - a class should solve a problem
3. Refactor regularly



# R6

More traditional OO Language approach but can produce code that is not "R-like"
Easier for OO programmers to pick up
Uses a different underlying mechanism than S3 and S4 (which are actually build on FP constructs)
Uses encapsulated functions instead

Main feature is the constructor method which has the same name of the class

```{r}
library(R6)

Accumulator <- R6Class("Accumulator", list(
  sum = 0,
  add = function(x = 1) {
    self$sum <- self$sum + x 
    invisible(self)
  })
)

A1 <- Accumulator$new()
A1$add()
A1$sum
A1$add(10)
A1$sum

```


Method Chaining

```{r}
A <-Accumulator$new()
A$add(10)$add(10)$sum
```
Initialize and Print

Initialize extends the constructor

```{r}
Dude <- R6Class("Dude", list(
  name = NULL,
  age = NA,
  initialize = function(name, age = NA) {
    stopifnot(is.character(name), length(name) == 1)
    stopifnot(is.numeric(age), length(age) == 1)
    
    self$name <- name
    self$age <- age
  }
))

Chuck <- Dude$new("Chuck", age = "thirty-eight")
Chuck <- Dude$new("Chuck", age = 38)
```
If you have more expensive validation requirements, implement them in a separate $validate() and only call when needed.

Print customizes printing

```{r}
Dude <- R6Class("Dude", list(
  name = NULL,
  age = NA,
  initialize = function(name, age = NA) {
    self$name <- name
    self$age <- age
  },
  print = function(...) {
    cat("Dude: \n")
    cat("  Name: ", self$name, "\n", sep = "")
    cat("  Age:  ", self$age, "\n", sep = "")
    invisible(self)
  }
))

Mikey <- Dude$new("Mikey")

```

